---
title: "Lab 1, Short Questions"
group: "Group 4"
geometry: margin=1in
output:
  pdf_document:
    toc: yes
    number_sections: yes
fontsize: 11pt
---

```{r load packages for homework 2, message=FALSE}
list.of.packages <- c("tidyverse", "patchwork", "GGally", "viridis", "hrbrthemes", "gridExtra", "ggthemes", "nnet", "MASS", "car")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(tidyverse)
library(patchwork)
library(GGally)
library(viridis)
library(hrbrthemes)
library(gridExtra) 
library(ggthemes)
## for multinomial log-linear models.
library(nnet)
## To use plor()
library(MASS)
## provide useful functions to facilitate the application and interpretation of regression analysis.
library(car)
```

# Strategic Placement of Products in Grocery Stores (5 points)

These questions are taken from Question 12 of chapter 3 of the textbook(Bilder and Loughin’s “Analysis of Categorical Data with R.

> *In order to maximize sales, items within grocery stores are strategically placed to draw customer attention. This exercise examines one type of item—breakfast cereal. Typically, in large grocery stores, boxes of cereal are placed on sets of shelves located on one side of the aisle. By placing particular boxes of cereals on specific shelves, grocery stores may better attract customers to them. To investigate this further, a random sample of size 10 was taken from each of four shelves at a Dillons grocery store in Manhattan, KS. These data are given in the *cereal_dillons.csv *file. The response variable is the shelf number, which is numbered from bottom (1) to top (4), and the explanatory variables are the sugar, fat, and sodium content of the cereals.*

```{r read cereal data, message=FALSE}
cereal <- read_csv('./cereal_dillons.csv')

```


## Recode Data 
(1 point) The explanatory variables need to be reformatted before proceeding further (sample code is provided in the textbook). First, divide each explanatory variable by its serving size to account for the different serving sizes among the cereals. Second, rescale each variable to be within 0 and 1. Construct side-by-side box plots with dot plots overlaid for each of the explanatory variables. Also, construct a parallel coordinates plot for the explanatory variables and the shelf number. Discuss whether possible content differences exist among the shelves.

```{r recode data}
# Set up the function for normalizing data between 0 and 1
stand01 <- function (x) {
  (x - min(x)) / ( max(x) - min(x)) 
  }

#Establish new dataframe
cereal2 <- data.frame(Shelf = cereal$Shelf, 
                      sugar = stand01(x = cereal$sugar_g / cereal$size_g), 
                      fat = stand01 (x = cereal$fat_g / cereal$size_g ), 
                      sodium = stand01 (x = cereal$sodium_mg / cereal$size_g ))

# Setting shelf up as a factor
cereal2 <- cereal2 %>%
  mutate(
    Shelf = factor(Shelf)
      )

#view(cereal2)

```


```{r boxplots}
# Use box plots to examine response variable with quantitative data
p1 <- cereal2 %>%
  ggplot(aes(x = Shelf, y = sugar, fill = Shelf)) + 
  geom_boxplot(varwidth = TRUE, alpha=0.2) +
  geom_jitter() +
  theme(plot.title = element_text(lineheight = 1, face = "bold")) +
  ggtitle("Sugar Boxplot") +
  ylab("Sugar Value (0-1)") + 
  xlab("Shelf Number")

p2 <- cereal2 %>%
  ggplot(aes(x = Shelf, y = fat, fill = Shelf)) + 
  geom_boxplot(varwidth = TRUE, alpha=0.2) +
  geom_jitter() +
  theme(plot.title = element_text(lineheight = 1, face = "bold")) +
  ggtitle("Fat Boxplot") +
  ylab("Fat Value (0-1)") + 
  xlab("Shelf Number")

p3 <- cereal2 %>%
  ggplot(aes(x = Shelf, y = sodium, fill = Shelf)) + 
  geom_boxplot(varwidth = TRUE, alpha=0.2) +
  geom_jitter() +
  theme(plot.title = element_text(lineheight = 1, face = "bold")) +
  ggtitle("Sodium Boxplot") +
  ylab("Sodium Value (0-1)") + 
  xlab("Shelf Number")

# Plot on 1 grid
grid.arrange(p1, p2, p3, nrow = 1, ncol = 3)

```
Observations: Cereal boxes placed on the second shelf appear to have the highest amount of sugar, ranging from 0.6 to 1.0 in standardized terms. The distribution of sugar content is much wider on the other shelves, going as low as 0.1.  Shelf 1 has the widest distribution of fat content. Shelves 1, 2, and 3 have a few cereal boxes with fat contents of 1.0 (in standardized terms). Shelf 4 has a maximum fat content of 0.6.  Shelf 1 has cereal boxes with the highest sodium content. Shelf 3 has outlier points in both directions (very low and very high sodium content). Shelf 2 and 4 have (relatively) tighter distributions of sodium content.

```{r parallel coordiantes plot}
p4 <- ggparcoord(cereal2, columns = 2:4, scale = "globalminmax", alphaLines = .4, groupColumn = 1,
    showPoints = FALSE, title = "Parellel Coordinate Plot: Shelf 1") +
  scale_color_manual(values=c("maroon","gray", "gray", "gray")) +
  theme(
    plot.title = element_text(size = 10) 
  )
      
p5 <- ggparcoord(cereal2, columns = 2:4, scale = "globalminmax", alphaLines = .4, groupColumn = 1,
    showPoints = FALSE, title = "Parellel Coordinate Plot: Shelf 2") +
  scale_color_manual(values=c("gray", "maroon", "gray", "gray")) +
  theme(
    plot.title = element_text(size = 10) 
  )

p6 <- ggparcoord(cereal2, columns = 2:4, scale = "globalminmax", alphaLines = .4, groupColumn = 1,
    showPoints = FALSE, title = "Parellel Coordinate Plot: Shelf 3") +
  scale_color_manual(values=c("gray", "gray", "maroon", "gray")) +
  theme(
    plot.title = element_text(size = 10) 
  )

p7 <- ggparcoord(cereal2, columns = 2:4, scale = "globalminmax", alphaLines = .4, groupColumn = 1,
    showPoints = FALSE, title = "Parellel Coordinate Plot: Shelf 4") +
  scale_color_manual(values=c("gray", "gray", "gray", "maroon")) +
  theme(
    plot.title = element_text(size = 10) 
  )

# Plot on 1 grid
grid.arrange(p4, p5, p6, p7, nrow = 2, ncol = 2)

```


shelf 1 has higher levels of sodium, while shelf 2 appears to have higher levels of sugar and a few observations with the highest fat content.  Shelf 3 has a large range of values for sugar, fat, and sodium.  Shelf 4 has low-to-medium normalized values for sugar, fat, and sodium. 

## Evaluate Ordinal vs. Categorical 
(1 point) The response has values of $1, 2, 3,$ and $4$. Explain under what setting would it be desirable to take into account ordinality, and whether you think that this setting occurs here. Then estimate a suitable multinomial regression model with linear forms of the sugar, fat, and sodium variables. Perform LRTs to examine the importance of each explanatory variable. Show that there are no significant interactions among the explanatory variables (including an interaction among all three variables).

Observation: The setting under which it would be desirable to take into account ordinality is a situation where 1 represents the lowest physical level of a shelf as well as the lowest preferred level by shoppers.  For example, a shelf one would be considered undersirable while a level of 4 would be desirable by all shoppers.  Although the dimensions of the shelves matter, it is typically the case the people prefer shelves 2 and 3 because they are easier to access relative to shelves 1 and 4. Under ideal conditions, Dillons would be able to order shelf levels by shopper preference (e.g. level 1 = really not desirable, level 2 = not desirable, level 3 = desirable, level 4 = really desirable).    

```{r multinomial logit regressions}
# Model with normalized explanatory variables
model_cereal_shelves_linear <- multinom(formula = Shelf ~ sugar + fat + sodium,
                                        data = cereal2)
summary(model_cereal_shelves_linear)

# Model with normalized explanatory variables + interactions terms
model_cereal_shelves_quadratic <- multinom(formula = Shelf ~ sugar + fat + sodium + 
                                             sodium:sugar + sodium:fat + sugar:fat,
                                           data = cereal2)

summary(model_cereal_shelves_quadratic)
```
For the block of coefficients in the models, the first row compares the coefficients for Shelf=2 to our baseline of Shelf=1. The second row compares the coefficients for Shelf=3 to our our baseline of Shelf=1. Finally, the third row compares the coefficients for Shelf=4 to our baseline of Shelf=1.   


```{r likelihood ratio tests: main effects}
# LRT for model with normalized explanatory variables only
lrt_cereal_main_effects <- Anova(model_cereal_shelves_linear)
lrt_cereal_main_effects
```

```{r likelihood ratio tests: interaction effects} 
# LRT for model with normalized explanatory variables + interaction terms
lrt_cereal_quadratic_effects <- Anova(model_cereal_shelves_quadratic)
lrt_cereal_quadratic_effects
```

Observation: We performed an LRT using Anova() to explore if a given explanatory variable $x_{r}$ is statistically significant over all response categories in our multinomial model (shelves: 1, 2, 3, 4). The null hypothesis is testing whether all coefficients in the odds ratios for a particular predictor are zero or not. 

Our null hypothesis is 

$$H_{0}: B_{2r} = B_{3r} = 0$$
While our alternative hypothesis is: 
$$H_{a}: B_{2r} \not = 0 or B_{3r} \not = 0$$

In both models the small p-values indicate that sodium and sugar are statistically significant and affect the probability of having the cereal being placed on a shelf level.  The p-value for the fat coefficient is quite large, meaning we fail to reject the null hypothesis that the fat coefficient is zero and the insufficient evidence indicates that fat content is not essential to our response variable. Similarly, the p-values for all interaction coefficients is quite large, meaning we also fail to reject the null hypothesis that all interactions are zero, meaning that the interaction terms are not essential to our response variable.



## Where do you think Apple Jacks will be placed? 
(1 point) Kellogg’s Apple Jacks (http://www.applejacks.com) is a cereal marketed toward children. For a serving size of $28$ grams, its sugar content is $12$ grams, fat content is $0.5$ grams, and sodium content is $130$ milligrams. Estimate the shelf probabilities for Apple Jacks.


```{r predictions for apple jacks}
# Set Apple Jack Variables: 
size <- 28
sugar <- 12/size
fat <- 0.5/size
sodium <- 130/size

# Re-scaling sugar to be within 0 and 1
sugar_all <- cereal$sugar_g / cereal$size_g
sugar <- (sugar - min(sugar_all)) / (max(sugar_all) - min(sugar_all))
# Re-scaling fat to be within 0 and 1
fat_all <- cereal$fat_g / cereal$size_g
fat <- (fat - min(fat_all)) / (max(fat_all) - min(fat_all))
# Re-scaling sodium to be within 0 and 1
sodium_all <- cereal$sodium_mg/cereal$size_g
sodium <- (sodium - min(sodium_all)) / (max(sodium_all) - min(sodium_all))

# Prediction dataframe
predict.data <- data.frame(sugar = sugar, fat = fat, sodium = sodium)
aj_shelf_probs <- predict(object = model_cereal_shelves_linear,
                          newdata = predict.data, type = "probs")
# Print results
aj_shelf_probs

```

Observation: The probability of Apple Jacks being placed on each shelf is: 

* Shelf 1: `r aj_shelf_probs[1]`\%
* Shelf 2: `r aj_shelf_probs[2]`\%
* Shelf 3: `r aj_shelf_probs[3]`\%
* Shelf 4: `r aj_shelf_probs[4]`\%

Therefore, Apple Jacks has the highest probability of being placed on shelf 2, followed by shelf 4, shelf 3 and shelf 1.  

## Figure 3.3 
(1 point) Construct a plot similar to Figure 3.3 where the estimated probability for a shelf is on the *y-axis* and the sugar content is on the *x-axis*. Use the mean overall fat and sodium content as the corresponding variable values in the model. Interpret the plot with respect to sugar content.

```{r create figure 3.3}

# Create prediction model to ensure probabilities are estiamted for all four shelves
predict.d <- data.frame(sodium = mean(cereal2$sodium), sugar = c(mean(cereal2$sugar), mean(cereal2$sugar)), fat = mean(cereal2$fat))
pi.hat <- predict(object = model_cereal_shelves_linear, newdata = predict.d, type = "probs")
#head(pi.hat)
#pi.hat[,1]

# Create plotting area first to make sure get the whole region with respect to x-axis
curve(expr = predict(object = model_cereal_shelves_linear, newdata = data.frame(sodium = mean(cereal2$sodium), 
                                                                                sugar = x,
                                                                                fat = mean(cereal2$fat)), type = "probs")[,1], 
      ylab = expression(hat(pi)), xlab = "Sugar", type = "n",
      xlim = c(min(cereal2$sugar), max(cereal2$sugar)),
      ylim = c(0,1), col = "#5F9EA0", lty = "solid", lwd = 2, n = 1000,
      panel.first = grid(col = "gray", lty = "dotted"))

# Create plots for all shelves
curve(expr = predict(object = model_cereal_shelves_linear, newdata = data.frame(sodium = mean(cereal2$sodium), 
                                                                                sugar = x,
                                                                                fat = mean(cereal2$fat)), type = "probs")[,1], 
      ylab = expression(hat(pi)), xlab = "Sugar",
      xlim = c(min(cereal2$sugar[cereal2$Shelf == 1]), max(cereal2$sugar[cereal2$Shelf == 1])),  
      ylim = c(0,1), col = "#5F9EA0", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = model_cereal_shelves_linear, newdata = data.frame(sodium = mean(cereal2$sodium), 
                                                                                sugar = x,
                                                                                fat = 
                                                                                  mean(cereal2$fat)), type = "probs")[,2], 
      ylab = expression(hat(pi)), xlab = "Sugar",
      xlim = c(min(cereal2$sugar[cereal2$Shelf == 2]), max(cereal2$sugar[cereal2$Shelf == 2])), 
      ylim = c(0,1), col = "#4682B4", lty = "longdash", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = model_cereal_shelves_linear, newdata = data.frame(sodium = mean(cereal2$sodium), 
                                                                                sugar = x,
                                                                                fat = mean(cereal2$fat)), type = "probs")[,3], 
      ylab = expression(hat(pi)), xlab = "Sugar",
      xlim = c(min(cereal2$sugar[cereal2$Shelf == 3]), max(cereal2$sugar[cereal2$Shelf == 3])), 
      ylim = c(0,1), col = "#6495ED", lty = "dotdash", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = model_cereal_shelves_linear, newdata = data.frame(sodium = mean(cereal2$sodium), 
                                                                                sugar = x,
                                                                                fat = mean(cereal2$fat)), type = "probs")[,4], 
      ylab = expression(hat(pi)), xlab = "Sugar",
      xlim = c(min(cereal2$sugar[cereal2$Shelf == 4]), max(cereal2$sugar[cereal2$Shelf == 4])), 
      ylim = c(0,1),col = "#00BFFF", lty = "dotted", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

legend(x = .01, y = 1, legend=c("Shelf 1", "Shelf 2", "Shelf 3", "Shelf 4"), lty=c("solid","longdash","dotdash", "dotted"),
       col=c("#5F9EA0","#4682B4","#6495ED", "#00BFFF"), bty="n", lwd = c(2,2,2,2))


```
The plot shows that the at low levels of sugar, there is higher probability that cereals will be placed on shelves 3 and 4.  However, at higher levels of sugar, there is a higher probability of the cereal being placed on shelf 2.  Finally, with increases to the the level of sugar, there is a slight increase in the probability that the cereal will be placed on shelf 1, although the probability starts to plateau at higher levels of sugar.  


## Odds ratios 
(1 point) Estimate odds ratios and calculate corresponding confidence intervals for each explanatory variable. Relate your interpretations back to the plots constructed for this exercise. 

```{r produce odds ratios}

# Find the standard deviation for each of the coefficients: sodium, sugar and fat: 
sd.cereal <- apply(X = cereal2[, -c(1)], MARGIN = 2, FUN = sd)
print("Standard Deviations of the Model Coefficients")
sd.cereal

# Compute confint for the coefficients for use in CI estimates
conf.beta <- confint(object = model_cereal_shelves_linear, level = 0.95)

# Odds ratios and CI
beta.hat2 <- coefficients(model_cereal_shelves_linear)[1, 2:4] # Shelf 2 Vs. Shelf 1
print("Shelf = 2 vs. Shelf = 1: Odds Ratios and CI")
round(exp(sd.cereal * beta.hat2), 4)
round(exp(sd.cereal * conf.beta [2:4 ,1:2 ,1]),5)

beta.hat3 <- coefficients(model_cereal_shelves_linear)[2, 2:4] # Shelf 1 Vs. Shelf 3
print("Shelf = 3 vs. Shelf = 1: Odds Ratio and CI")
round(exp(sd.cereal * beta.hat3 ), 4)
round(exp(sd.cereal * conf.beta [2:4 ,1:2 ,2]), 5)

beta.hat4 <- coefficients(model_cereal_shelves_linear)[3, 2:4] # Shelf 1 Vs. Shelf 4
print("Shelf = 4 vs. Shelf = 1: Odds Ratio and CI")
round(exp(sd.cereal * beta.hat4 ), 4)
round(exp(sd.cereal * conf.beta [2:4 ,1:2 ,3]), 5)

```
Observations: 

**Shelf 2 Vs. Shelf 1:** 
The estimated odds of a cereal being on shelf 2 vs. shelf 1 change by 2.06 times for a 0.27 increase in sugar holding the other variables constant.  With 95% confidence, the odds of level 2 instead of level 1 change by 0.14 to 29.68 times when sugar changes by 0.27, while holding the other variables constant.  The estimated odds of a cereal being on shelf 2 vs. shelf 1 change by 3.37 times for a 0.30 (rounded) increase in fat holding the other variables constant. With 95% confidence, the odds of level 2 instead of level 1 change by 0.87 to 13.04 times when fat changes by 0.30, while holding the other variables constant. The estimated odds of a cereal being on shelf 2 vs. shelf 1 change by 0.02 times for a 0.23 increase in sodium holding the other variables constant. With 95% confidence, the odds of shelf 2 instead of shelf 1 change by 0.001 to 0.439 times when sodium changes by 0.23, while holding the other variables constant.  

Relationship to previous plots: Figure 3 shows that probability of being on shelf 2 increases with higher levels of sugar, which supports the odds ratio of 2.06. Both sugar and fat have odds ratios greater then 1, which makes sense when assessing the parallel coordinate plots for shelf 1 and shelf 2 as it shows that cereals on shelf 2 generally have higher sugar values and mid-to-higher fat values relative to shelf 1.  Similarly, the the boxplots for sugar and fat demonstrate higher values relative to shelf 1. An odds ratio less than 1 for sodium (0.02) reflects the findings in the parallel coordinate plot and the box plots as shelf 1 has higher sodium values relative to shelf 2.      

**Shelf 3 Vs. Shelf 1:** 
The estimated odds of a cereal being on shelf 3 vs. shelf 1 change by 0.04 times for a 0.27 increase in sugar holding the other variables constant. With 95% confidence, the odds of level 3 instead of level 1 change by 0.0028 to 0.49 times when sugar changes by 0.27, while holding the other variables constant. The estimated odds of a cereal being on shelf 3 vs. shelf 1 change by 0.85 times for a 0.30 (rounded) increase in fat holding the other variables constant. With 95% confidence, the odds of level 3 instead of level 1 change by 0.21 to 3.49 times when fat changes by 0.30, while holding the other variables constant.  The estimated odds of a cereal being on shelf 3 vs. shelf 1 change by 0.003 times for a 0.23 increase in sodium holding the other variables constant. With 95% confidence, the odds of shelf 3 instead of shelf 1 change by 0.0001 to 0.1223 times when sodium changes by 0.23, while holding the other variables constant.  

Relationship to previous plots: Figure 3 shows that probability of being on shelf 3 decreases with higher levels of sugar, which supports the odds ratio less than 1. In fact, sugar and sodium have odds ratios that are only 0.04 and 0.003, respectively.  The difference between the variables in shelf 3 versus shelf 1 are best depicted in the boxplots for these variables as shelf 1 has higher sugar and sodium values relative to shelf 3.  This is also apparent in the parallel coordinate plot that shows the difference between sodium values for shelf 1 versus sodium values for shelf 3.  For fat, which has an odds ratio of 0.85, it is more difficult to tell the relative difference for all of the plots.      


**Shelf 4 Vs. Shelf 1:** 
The estimated odds of a cereal being on shelf 4 vs. shelf 1 change by 0.05 times for a 0.27 increase in sugar holding the other variables constant. With 95% confidence, the odds of level 4 instead of level 1 change by 0.0036 to 0.6084 times when sugar changes by 0.27, while holding the other variables constant. The estimated odds of a cereal being on shelf 4 vs. shelf 1 change by 0.77 times for a 0.30 (rounded) increase in fat holding the other variables constant. With 95% confidence, the odds of level 3 instead of level 1 change by 0.19 to 3.16 times when fat changes by 0.30, while holding the other variables constant. The estimated odds of a cereal being on shelf 4 vs. shelf 1 change by 0.003 times for a 0.23 increase in sodium holding the other variables constant. With 95% confidence, the odds of shelf 4 instead of shelf 1 change by 0.0001 to 0.1301 times when sodium changes by 0.23, while holding the other variables constant. 

Relationship to previous plots: Figure 3 shows that probability of being on shelf 4 decreases with higher levels of sugar, which supports the odds ratio less than 1. When comparing shelf 4 to shelf 1, sugar and sodium have odds ratios that are only 0.05 and 0.003, respectively.  The difference between sodium on the two shelves is apparent in both the sodium boxplot and the parallel coordinate plots for shelf 1 and shelf 4. The difference between sugar on both the parallel coordinate plot and the boxplot is less pronounced but there are difference between the median values in the boxplot. Fat, which has an odds ratio of 0.77, is presented best in the parallel coordinate plots, as the fat values seem to be higher on shelf 1 relative to shelf 4.     


# Alcohol, self-esteem and negative relationship interactions (5 points)

Read the example **'Alcohol Consumption'** in chapter 4.2.2 of the textbook(Bilder and Loughin’s “Analysis of Categorical Data with R). This is based on a study in which moderate-to-heavy drinkers (defined as at least 12 alcoholic drinks/week for women, 15 for men) were recruited to keep a daily record of each drink that they consumed over a 30-day study period. Participants also completed a variety of rating scales covering daily events in their lives and items related to self-esteem. The data are given in the *DeHartSimplified.csv *data set. Questions 24-26 of chapter 3 of the textbook also relate to this data set and give definitions of its variables: the number of drinks consumed (`numall`), positive romantic-relationship events (`prel`), negative romantic-relationship events (`nrel`), age (`age`), trait (long-term) self-esteem (`rosn`), state (short-term) self-esteem (`state`).

The researchers stated the following hypothesis:

> *We hypothesized that negative interactions with romantic partners would be associated with alcohol consumption (and an increased desire to drink). We predicted that people with low trait self-esteem would drink more on days they experienced more negative relationship interactions compared with days during which they experienced fewer negative relationship interactions. The relation between drinking and negative relationship interactions should not be evident for individuals with high trait self-esteem.*

```{r read drinking data, message=FALSE}
drinks <- read.csv('DeHartSimplified.csv')
```

## EDA 
(2 points) Conduct a thorough EDA of the data set, giving special attention to the relationships relevant to the researchers' hypotheses. Address the reasons for limiting the study to observations from only one day.

```{r drinking EDA}

```

> 'Fill this in: What do you learn?' 

## Hypothesis One 
(2 points) The researchers hypothesize that negative interactions with romantic partners would be associated with alcohol consumption and an increased desire to drink. Using appropriate models, evaluate the evidence that negative relationship interactions are associated with higher alcohol consumption and an increased desire to drink. 

> 'Fill this in: What do you learn?' 

## Hypothesis Two
(1 point) The researchers hypothesize that the relation between drinking and negative relationship interactions should not be evident for individuals with high trait self-esteem. Conduct an analysis to address this hypothesis.

> 'Fill this in: What do you learn?' 